---
title: "Supply and Demand for the United States Urogynecology Workforce, 2025-2050: Moving Beyond the Head Count"
author:
- name: Tyler Muffly, MD
  affiliation: Department of Obstetrics and Gynecology, Denver Health Medical Center,
    University of Colorado School of Medicine
  email: tyler.muffly@dhha.org
  corresponding: true
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    fig_caption: true
    fig_height: 6
    fig_width: 8
    reference_docx: templates/green_journal_reference.docx
running_head: Urogynecology Supply and Demand
precis: Projected national urogynecology supply grows faster than demographic
  demand, but aggregate head count does not establish geographic access or
  operative capacity.
word_count: 3,700 words
abstract_word_count: 300 words
conflict_of_interest: The authors report no conflict of interest.
funding: "This research received no specific grant from any funding agency in the
  public, commercial, or not-for-profit sectors."
corresponding_author: 'Tyler Muffly, MD, Department of Obstetrics and Gynecology,
  Denver Health Medical Center, 777 Bannock Street, Denver, CO 80204. Email: tyler.muffly@dhha.org'
keywords:
- urogynecology
- pelvic floor disorders
- physician workforce projection
- workforce supply and demand
- clinical practice departure
- fellowship training
bibliography: references_WORKFORCE_CLIFF.bib
csl: csl/american-medical-association.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Core packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(knitr)
  library(kableExtra)
  library(scales)
})

# Load statistics + figure functions
source(here("manuscript/R/workforce_statistics.R"))
source(here("manuscript/R/workforce_figures.R"))
source(here("manuscript/R/create_figure_stock_flow_design.R"))

# Knitr options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/",
  dpi = 300
)
```

## Abstract

**OBJECTIVE:** To project the United States urogynecology and reconstructive pelvic surgery (URPS) workforce supply-and-demand balance through 2050, integrating an empirically estimated clinical-practice departure hazard with demographic demand for pelvic floor disorder care.

**METHODS:** We assembled the national both-pathway URPS workforce from American Board of Obstetrics and Gynecology and American Board of Urology certification records linked to administrative data, and classified clinical-practice departure from six administrative streams. A pooled age-band departure hazard, estimated over the fully observable 2016-2021 window with gynecologic oncology as comparator, drove a dynamic age-structured supply projection adding annual ACGME fellowship completers, with supply also expressed as productivity-adjusted capacity. Demand combined the projected United States population of women aged 65 years or older (US Census Bureau 2023) and those with pelvic floor disorders. Uncertainty used Monte Carlo simulation and multidimensional sensitivity analyses.

**RESULTS:** The active both-pathway URPS workforce comprised `r get_baseline("URPS")` urogynecologists; `r get_departure_events("URPS")` departures over 2016-2021 implied a weighted annual departure rate of `r get_annual_rate("URPS")`%. Annual fellowship output (`r get_annual_entrants("URPS")`) exceeded near-term departures, and supply grew about 79% by 2050 (index 179, 2025 = 100). Meanwhile women aged 65 years or older rose 29% (34.7 to 44.8 million) and those with pelvic floor disorders 19% (33.1 to 39.5 million), so supply outpaced demand: urogynecologists per 100,000 older women rose from 3.9 to 5.4, and women with a pelvic floor disorder per urogynecologist fell from roughly 24,800 to 16,500. The departure endpoint is provisional pending physician-level adjudication; demand is demographic rather than utilization-derived; and operative and geographic capacity are unresolved.

**CONCLUSIONS:** Integrating an empirically estimated departure hazard with demographic demand, projected United States urogynecology supply grows faster than demand through 2050. These national aggregate findings rest on a provisional classification and do not establish geographic access, operative capacity, or local-market adequacy.

# Introduction

Urogynecology and reconstructive pelvic surgery (URPS) is a board-certified subspecialty providing advanced surgical care for pelvic organ prolapse, urinary incontinence, and other pelvic floor disorders, conditions whose prevalence increases with age and that affect nearly one in four United States women.[@nygaard2008] Because pelvic floor disorders concentrate in older women, demand for urogynecologic care is coupled to demographic aging: the United States population of women aged 65 years or older is projected to grow by roughly one-third between 2025 and 2050. Whether the workforce can meet that rising demand depends jointly on how many physicians enter subspecialty practice through fellowship training and how many leave clinical practice as a small, geographically concentrated, and aging cohort retires. As a comparator subspecialty with a similar age structure and training pathway, we retain board-certified gynecologic oncology, which provides definitive surgical treatment for gynecologic malignancies.[@siegel2023] Because our demand model captures pelvic floor disorder prevalence and not cancer incidence, gynecologic oncology enters this analysis on the supply side only, as a replacement comparator and as the comparator cohort that stabilizes the pooled departure hazard; we model demand, and therefore supply relative to demand, for urogynecology alone. A voluntary focused-practice designation in minimally invasive gynecologic surgery (MIGS) is examined separately as an exploratory cohort (Appendix S13), because it is designation-based rather than board-certified and its inflow and departure ascertainment are not comparable.[@abog_fpmigs; @vargas2018]

As large cohorts approach retirement, departures may soon outpace the new subspecialists that fellowship programs produce. Worsening geographic access to gynecologic oncologists has been documented,[@desjardins2023; @ricci2017; @boland2024] and broader analyses anticipate a mismatch between an aging supply and rising demand across surgical subspecialties.[@dall2013; @adashi2025; @sharma2024] Existing assessments, however, typically infer departure from a single administrative signal such as Medicare billing cessation, which can misclassify low-volume subspecialists and those shifting to nonclinical roles, and rely on demographic aging without quantifying uncertainty in the timing of individual exits. Prior surgical workforce models nonetheless inform our design: demand-side models apply age-specific procedure-use rates to population projections;[@etzioni2003; @edwards2014] dynamic stock-and-flow models age cohorts forward and can express capacity as full-time equivalents rather than head count;[@fraher2013] good-practice standards emphasize explicit assumptions and quantified uncertainty;[@leejt2024] and contemporary projections combining certifications, completions, and retirement find training expansion alone insufficient to offset losses.[@ellison2018] Federal supply-demand modeling cannot separate the gynecologic subspecialties from generalist obstetrician-gynecologists, leaving subspecialty replacement unmeasured.[@rayburn2025complexities] A single prior national projection joined urogynecology supply and demand, comparing a projected subspecialist head count with the projected number of women with pelvic floor disorders and finding a large, persistent burden per subspecialist, but it assumed retirement rather than estimating departure and counted physicians rather than clinical capacity.[@brueseke2016] Because a head count alone can mislead, since it is clinical capacity, not the number of bodies, that patients experience,[@stitzenberg2015; @temkin2022fte] we build on that projection by estimating departure empirically from six administrative streams, counting an active both-pathway roster validated by billing, expressing supply as productivity-adjusted capacity, and quantifying uncertainty. Whether the anticipated shortage is a near-term problem of aggregate supply rather than of the distribution of a growing but geographically concentrated workforce has not been tested against actual graduate output and explicit demand; we provide that test.

To address these gaps, we assembled the national United States both-pathway urogynecology workforce (the obstetrics-gynecology and urology certifying pathways), with gynecologic oncology (GO) as a comparator, restricted each to physicians in active clinical practice, and classified clinical-practice departure using a framework that integrated six administrative evidence streams. On the supply side, we projected each active workforce with a dynamic age-structured model driven by an empirically estimated age-band departure hazard and ACGME fellowship completions, and expressed supply as productivity-adjusted capacity as well as head count. On the demand side, we projected the population of older United States women and the population with pelvic floor disorders through 2050, and combined supply with demand into per-capita coverage and a capacity-to-demand index. This analysis addresses national aggregate supply and demand; it does not resolve geographic distribution, effective operative capacity, or whether aggregate adequacy is realized in any local market. We hypothesized that recent fellowship output would replace near-term departures but that demographic growth would reduce per-capita coverage.

# Materials and Methods

## Study Design

This was a dynamic stock-flow simulation projection built from retrospective administrative trend data, following the supply-forecast design used for other subspecialty workforces.[@shah2026] We estimated clinical-practice departures over 2016-2021 and used those estimates in a dynamic age-structured projection through 2029 of an estimated 2025 active baseline derived from the most recently available administrative signals (Medicare through 2023, Open Payments through 2024), forecasting workforce entrants (fellowship graduates) and workforce exit (empirical departures) separately and combining them each year. Because the source data predate 2025, the baseline is an estimate carried forward from the latest signals rather than a directly observed 2025 census, and could misclassify physicians who departed in 2023-2024 or omit those entering after the data freeze (baseline-lag sensitivity, Appendix S1). Analyses drew on public-record professional data (board certification, NPPES, Medicare administrative files); the Denver Health institutional review board determined that the study did not constitute human-subjects research. The study followed STROBE reporting guidelines for observational research.[@vonelm2007]

## Data Sources

### Physician Population Identification

Physicians in Gynecologic Oncology and Urogynecology and Reconstructive Pelvic Surgery were identified from American Board of Obstetrics and Gynecology (ABOG) publicly available certification records; board certification, rather than self-reported taxonomy codes in administrative databases, defined subspecialty. Because urogynecology and reconstructive pelvic surgery is certified through both ABOG (obstetrics-gynecology pathway) and the American Board of Urology (ABU, urology pathway), and the ACGME fellowship graduate count spans both, we made the URPS baseline a consistent both-pathway population: the ABOG-certified urogynecologists plus the ABU-certified urogynecologists not already included. ABU urogynecologists were matched to National Provider Identifiers and de-duplicated against the ABOG cohort. An initial de-duplication against the full ABOG certification file removed 85 ABU-certified urogynecologists whose National Provider Identifier matched a same-named ABOG *generalist* rather than an ABOG urogynecologist; on adjudication, 38 of these (28 carrying the Female Pelvic Medicine and Reconstructive Surgery taxonomy on their own National Provider Identifier, and 10 with a demonstrable false name match to the ABOG record) were reinstated as genuine urology-pathway urogynecologists, while 47 coded only with a general-urology taxonomy were held out pending confirmation. This yielded 308 net-new ABU urogynecologists in the identified roster (270 net-new plus 38 reinstated dual-boarded); the 279 whose urogynecology subspecialty certification was effective by 2023 enter the both-pathway 2023 board-certified active baseline. For the age-structured projection, each physician's age was reconstructed from the urogynecology subspecialty certification year where recorded (981 ABOG physicians, from the American Board of Obstetrics and Gynecology sub-certification start date) and otherwise from a fellowship-timing proxy (46 ABOG physicians as primary certification plus three years, and all 279 ABU physicians as primary certification plus two years), so that every active physician is carried in the stock and exposed to age-specific departure risk. The `r get_baseline("URPS")` both-pathway baseline therefore comprises the 1,027 ABOG-certified active urogynecologists plus these 279 ABU physicians.[@plaska2025] As direct evidence that these urology-pathway physicians actively practice urogynecology rather than holding certification nominally, 87% billed at least one procedure from the empirically defined female pelvic medicine and reconstructive surgery code set in Medicare fee-for-service in 2024 (the most recent year available; median 666 services per physician), and 44 of them performed sling operations for stress urinary incontinence (Current Procedural Terminology code 57288), at least 851 such operations in aggregate; because Medicare fee-for-service excludes commercial, Medicaid, and Medicare Advantage claims and suppresses provider-code cells below eleven services, these counts are conservative lower bounds. Minimally Invasive Gynecologic Surgery is a voluntary ABOG Focused Practice Designation rather than a subspecialty board certification; it is analyzed separately as an exploratory designation-based cohort (below) and is not pooled with the two board-certified subspecialties. No geographic restriction was applied at any stage: both the physician denominator (the active board-certified stock and its departures) and the ACGME fellowship-completion numerator are national United States counts, so the completion-to-departure ratio compares geographically like with like (the two cohorts include fewer than three subspecialists outside the contiguous states, all in Hawaii).

National Provider Identifier (NPI) records from the National Plan and Provider Enumeration System (NPPES) provided longitudinal practice locations and demographic information.[@nber2024] We supplemented these data with Medicare Part B claims (2013-2023), Medicare Part D prescriber files (2013-2023), and CMS Open Payments records (2013-2024), the most recent public releases at the data-freeze date, to establish clinical activity patterns and identify departures from practice. Because confirming a sustained multi-year absence requires follow-up beyond the last active year, these freeze dates bound which departure years can be fully adjudicated (see Departure Rate).

### Clinical-Practice Departure Detection

We employed a hierarchical six-source departure detection framework integrating administrative evidence streams, each ranked by priority and weighted by reliability (Appendix Table S2, Supplemental Digital Content): a lapse in American Board of Medical Specialties (ABMS) Maintenance of Certification (MOC);[@abms2024] permanent NPPES deactivation without reactivation; cessation of Open Payments industry payments for at least two years; disenrollment from Medicare provider enrollment in PECOS; and cessation of Medicare Part D prescribing or Part B claims for at least three years. When signals conflicted, the framework resolved them by weighted consensus under the priority order. Because cessation of industry-payment reporting is not a validated indicator of leaving clinical practice, the primary departure definition requires corroboration by at least one non-Open-Payments anchor, defined as Medicare Part B cessation, Medicare Part D cessation, permanent NPPES deactivation, or ABMS certification lapse; an Open Payments cessation alone does not classify a departure, and PECOS disenrollment is a detection signal that does not by itself anchor the primary endpoint. The prior definition that allowed Open Payments-only departures is retained as a sensitivity (Appendix S15).

The primary endpoint is governed by a single uniform cessation rule applied to both cohorts: a departure required a sustained three-year absence of the qualifying (anchoring) activity signals, namely Medicare Part B claims and Part D prescribing; no subspecialty-specific inactivity threshold was used (Supplemental Digital Content, Appendix S4). The shorter two-year threshold noted above applies only to the Open Payments detection signal, which is supporting-only and cannot by itself classify a departure, so it does not set the three-year rule that defines the primary endpoint.

Because these signals capture cessation of clinical activity rather than a reason for leaving, the outcome is **sustained departure from active clinical practice** (a multi-year absence of activity signals through the available follow-up, not necessarily lifelong permanence), each classified departure represents an outflow from the active subspecialty headcount regardless of cause. It encompasses age-related retirement together with other sustained exits (career change, relocation abroad, non-clinical work), and administrative data cannot distinguish among them.[@rotenstein2025; @elemosho2026; @gettel2023] Studies of physicians leaving individual practices find many exits are moves or role changes rather than retirements, though leaving one employer is not equivalent to leaving national clinical practice.[@oconnell2023] We therefore report clinical-practice departure throughout and reserve "retirement" for age-related exits. Because the physician-level adjudication of this classification is not yet complete (Appendix S7), we present departure classification as a provisional administrative model rather than a validated measurement of departure; the high completion-to-departure ratios make the headcount conclusion robust to substantial misclassification: the headcount balance would fall below one-for-one only if true departures reached the entrant count. Conditional on the current classifier, active baseline, pooled-hazard model, and the assumption that all completers enter practice, roughly `r get_tipping_missed_range(1.0)` of true departures would have to be missed to overturn the conclusion; under a 70% graduate-to-practice conversion the corresponding tipping points are roughly `r get_tipping_missed_range(WORKFORCE_CONVERSION_FLOOR)`. This single-axis tipping point is one component of a broader uncertainty analysis, not a substitute for validating the endpoint: because several moderate biases (missed deaths, missed non-Medicare departures, lower graduate-to-practice conversion, reduced entrant clinical effort, baseline over-counting, and specialty hazard differences) act in the same direction, they could compound, so it should be read alongside the structural sensitivity envelope rather than as evidence that classifier performance is unimportant. Missed departures, not false-positive departures, are the error that could threaten the result, and the departure counts await individual validation.

### Forecast of Workforce Entrants (Fellowship Pipeline)

Annual potential entrants were measured as fellowship completers (graduates), the number of physicians finishing training each year and thus potentially available to enter practice, rather than positions filled at the start of training. Graduation does not guarantee that an individual enters U.S. clinical practice in the subspecialty, becomes board-certified, or does so within the modeled year; the 70% to 100% practice-conversion sensitivity brackets this gap between completion and observed active practice. For Gynecologic Oncology and URPS we used the ACGME Data Resource Book count of graduating residents (Table D.5), averaged over the four most recent published academic years (2020-2021 through 2023-2024):[@acgme_drb] a mean of `r get_annual_entrants("GO")` per year for Gynecologic Oncology (range 70 to 79) and `r get_annual_entrants("URPS")` for URPS (range 61 to 66). Because URPS is certified and trained through both the obstetrics-gynecology and urology pathways, the URPS graduate count sums both sponsoring pathways in Table D.5 (obstetrics-gynecology-sponsored 45 to 51 plus urology-sponsored 15 to 16 per year), matching the both-pathway active stock; the obstetrics-gynecology pathway alone would have undercounted graduates by about a quarter and mismatched the denominator. Using a multi-year mean rather than a single year guards against year-to-year fluctuation and feeds the uncertainty analysis.

We retained the NRMP Specialties Matching Service certified-position counts (`r get_nrmp_entrants("GO")` and `r get_nrmp_entrants("URPS")` for Gynecologic Oncology and URPS in the 2026 appointment year, the most recent match; 87 and 70 were filled; the URPS match is a both-pathway urology and obstetrics-gynecology track) as an upper program-capacity benchmark inflow.[@nrmp2025] Because NRMP counts fellows beginning training, who enter practice only after completing their fellowship, this benchmark is applied to the projection with a pathway-specific completion lag (the urology URPS fellowship is 2 years, the obstetrics-gynecology URPS and Gynecologic Oncology fellowships 3 years), so a matched cohort raises graduate output only in its expected completion year. Filled positions modestly exceed graduates because these fellowships have grown over the past decade,[@pyskir2022; @vargas2018] so ACGME completers are the appropriate measure of training output for the completion-to-departure flow analysis, representing potential rather than confirmed active-practice entrants. Training attrition is small (about 97% completion). We assumed stable graduate output over the horizon; the stability of that assumption, given recent applicant-market trends, is examined in the Discussion.[@humphries2026]

### Operative-Workforce Analysis (Exploratory)

Because a surgeon can leave operative practice while remaining clinically active, we performed an exploratory secondary analysis of operative-workforce replacement for Gynecologic Oncology and URPS, using Medicare Part B by-service claims (procedure-level HCPCS codes per physician per year) to identify sustained cessation of Medicare-observed qualifying surgery among established operators (at least three qualifying major operations per year in two of three preceding years), operationalized as sustained absence of Medicare-observed qualifying surgery.[@poushay2018; @conn2018] We estimated graduate operative entry with time-to-event methods, computed operator-count and procedure-volume-weighted replacement, and assessed the concordance of each apparent operative cessation with independent signals of complete clinical-practice departure (total Medicare billing cessation and NPPES deactivation). Because Medicare fee-for-service captures only the older-patient share of operative volume and cannot observe commercial or facility-billed surgery, all operative quantities are Medicare-provisional and this analysis is reported chiefly as a validity assessment; definitions, thresholds, time-to-event and volume-weighted estimates, concordance and validity analyses, and sensitivity analyses are in the Supplemental Digital Content (Appendix S11).

## Workforce Projection

### Forecast of Workforce Exit (Departure Rate)

Age-band departure hazards were derived empirically from the study cohort's own departure history rather than imported from a general-physician model, as other recent subspecialty workforce projections have done (which combine Social Security Administration mortality with Health Resources and Services Administration retirement probabilities).[@shah2026] Using the multi-source classification, we built a person-year life table in which each physician contributed one person-year per active year and a departure event in the year of departure, with age reconstructed from the board-certification year. Because departure events are sparse within any single subspecialty age band (`r get_departure_events("GO")` departure events in Gynecologic Oncology and `r get_departure_events("URPS")` in both-pathway URPS, `r get_departure_events()` pooled, over the 2016-2021 window; Appendix S6b reports the event and person-year counts by band), we estimated the primary age-band hazard by pooling the two board-certified cohorts, gynecologic oncology and both-pathway urogynecology, into a single shared age-band hazard, the parsimonious choice given only two cohorts and sparse events. We also fit a hierarchical partial-pooling model (a shared age-band shape plus a penalized specialty deviation, fit by penalized maximum likelihood with a weakly informative prior on the between-specialty variance; the frequentist analog of a Bayesian partial pool, and with only two cohorts an unpenalized fit collapses to full pooling) and a fully unpooled model, both reported as bracketing sensitivities (Appendix S6b, Appendix Table S28). The exploratory MIGS cohort differs in ascertainment, certification status, and graduate numerator, so its departure events are excluded from the primary hazard and it is projected with the fully pooled hazard; a MIGS-inclusive hazard is reported as a sensitivity in Appendix S6b. We applied the pooled hazard to each cohort's own 2025 active-age distribution. Because both cohorts share the pooled hazard, the reported per-subspecialty departure rates differ only through each cohort's age structure; the small between-specialty deviation estimated in the partial-pooling sensitivity confirms the two rates stay close. Appendix S6b reports the hierarchical partial-pooled and fully unpooled subspecialty-specific hazards as sensitivities that bracket the primary pooled estimate. To avoid right-censoring, hazards were estimated only over the fully observable window 2016-2021: because the classifier requires a sustained multi-year absence of billing and enrollment signals and the underlying data end in 2023-2024, departures dated 2022-2023 cannot yet be confirmed under the three-year administrative follow-up rule and accumulate as provisional over-flags, so they were excluded from hazard estimation and treated as right-censored. The `r primary_hazard_adjective()` age-band hazard applied to each subspecialty's estimated 2025 active-age distribution (carried forward from the latest administrative signals) implies weighted departure rates of `r get_annual_rate("GO")`% for Gynecologic Oncology and `r get_annual_rate("URPS")`% for URPS. These model-implied weighted rates were numerically lower than the 1.5% to 1.7% per year national all-surgeon attrition estimate,[@elemosho2026] although direct comparison is limited by differences in cohort definition, administrative ascertainment, and hazard modeling (the published estimate placed gynecologic oncology among the lowest-attrition specialties, adjusted hazard ratio 0.73 versus general surgery). We report the full-window (2016-2023) and further-restricted (2016-2019) hazards as sensitivities. In the projection these age-band hazards, not a single fixed rate, are reapplied each year to the cohort as it ages.

### Projection Model

We projected each active workforce with a dynamic age-structured stock-flow model over the four-year horizon (2025-2029), following the supply-forecast framework applied to other subspecialties[@shah2026] but replacing imported general-physician attrition with empirically derived age-band hazards (Departure Rate): each year the cohort ages, age-band hazards yield departures, and graduates enter at their entry age (full specification and the static single-rate benchmark, Appendix S1). Because the age-band hazards are estimated over 2016-2021 and the same window's cohort survival is reproduced, this is an in-sample calibration check rather than out-of-sample temporal validation. On that basis the model reproduced observed active headcount to within about 1% over a two-to-five-year horizon (Appendix Table S4), and Stratifying the back-test by age band (the 2016 cohort's survival to 2021 within each band) showed calibration to within about `r get_backtest_band_maxerr(50)`% in every well-populated age band (those holding the large majority of physicians); larger percentage errors were confined to the sparsely populated oldest bands, where as few as four to nine physicians per cohort make percentage errors on small counts uninformative (Appendix Table S31). The historical back-test cohort is the ABOG-pathway population; ABU net-new urogynecologists have no pre-2024 roster and are not in it.

```{r figure-stock-flow-design, fig.cap="**Figure 1.** Design of the dynamic age-structured stock-flow supply forecast. The active workforce is carried as an age-structured stock, initialized at the estimated 2025 baseline (Gynecologic Oncology single-pathway; both-pathway URPS). Each projection year the stock ages one year, the empirical pooled age-band departure hazard (estimated over the fully observable 2016-2021 window from Gynecologic Oncology and both-pathway URPS) removes departures, and annual ACGME fellowship completers enter at their entry age; entrants and departures are combined each year across the 2025-2029 horizon (extended to 2050 to align with demographic demand). Parameter uncertainty is propagated by Monte Carlo simulation and a multidimensional sensitivity grid, yielding conservative, status-quo (primary), and optimistic scenarios. Model outputs are the annual headcount trajectory, the fellowship-completion-to-departure ratio, productivity-adjusted full-time-equivalent capacity, and supply relative to demand. URPS, Urogynecology and Reconstructive Pelvic Surgery; GO, Gynecologic Oncology; ACGME, Accreditation Council for Graduate Medical Education; FTE, full-time equivalent.", fig.width=9.2, fig.height=5.0}
print(fig_stock_flow_design())
```

### Uncertainty

We characterized uncertainty in the 2029 count with a `r format(WORKFORCE_MONTE_CARLO_ITERATIONS, big.mark = ",")`-iteration Monte Carlo parameter-uncertainty simulation (random seed 20260718) drawing each age-band hazard from its Beta(events + 0.5, person-years - events + 0.5) posterior and each subspecialty's annual graduate output by empirical resampling of the four observed annual completer counts (no distributional assumption); the 2.5th and 97.5th percentiles are the reported interval. Because it reflects parameter uncertainty around the expected workforce rather than individual stochastic events and omits baseline-classification error, it is a partial parameter-uncertainty interval (it also omits ABU-roster, graduate-conversion, model-form, death-ascertainment, baseline-definition, physician-level classification, and MIGS-pooling uncertainty), not a full predictive interval, so the reported bands understate total uncertainty and the scenario ranges are a fuller expression of it than these narrow Monte Carlo bands. A multidimensional sensitivity analysis, passed through the same model, crossed the observation window (2016-2021, 2016-2019, 2016-2023), the departure-rate multiplier (up to threefold, spanning the general-physician rates of the HRSA Health Workforce Simulation Model), and graduate practice-conversion (70% to 100%; Appendix S12).[@gettel2023] Following the fellowship-inflow scenario approach of recent subspecialty analyses,[@holtzman2025; @shah2026] the primary projection is the status-quo entrant scenario (recent four-year graduate mean), bracketed by conservative (70% conversion) and optimistic (NRMP benchmark) scenarios.

## Fellowship-Completion-to-Departure Ratio Calculation

The fellowship-completion-to-departure ratio measures **headcount flow balance**, the ratio of annual fellowship completions to mean annual departures. Because the numerator counts fellowship completions rather than verified entrants into active subspecialty practice, the ratio is a flow-balance measure, not a count of confirmed active replacements; the graduate-to-practice conversion sensitivity (70% to 100%) addresses this gap directly.

$$\text{Fellowship-Completion-to-Departure Ratio} = \frac{\text{Annual Fellowship Completions}}{\text{Mean Annual Departures}}$$

A ratio above 1.0 means fellowship completions exceed classified departures, 1.0 exact numerical balance, and below 1.0 that classified departures exceed fellowship completions; we apply a narrow +/-5% band (above replacement >1.05, at replacement 0.95 to 1.05, below replacement <0.95) only to avoid over-reading ratios near unity, which does not arise here because every ratio is far from 1.0. These describe whether new graduate headcount offsets departing headcount; they do not measure whether the workforce meets population demand, geographic access, or clinical capacity. A 1.20 threshold is reported as a labeled policy buffer, not a measure of adequacy. One new graduate is not assumed to replace the clinical capacity of one established surgeon. Because this ratio is a supply-side flow balance, it is paired below with an explicit demand denominator.

## Demand Estimation and Capacity Relative to Demographic Demand

To situate the projected workforce against the population it serves, we estimated demographic demand for urogynecologic care from 2025 through 2050. The population of United States women aged 65 years or older, the age group in which pelvic floor disorders concentrate, was taken from the US Census Bureau 2023 National Population Projections; the population living with a pelvic floor disorder was estimated by applying published age-specific pelvic floor disorder prevalence to those projections.[@nygaard2008] Because these demand estimates are demographic rather than utilization-derived, they represent population need under stable prevalence and care-seeking rather than observed procedure volume; a utilization-based demand model is a planned extension. We modeled demand for urogynecology only. Gynecologic-oncology demand is driven by cancer incidence, which the pelvic floor disorder prevalence model does not capture, so we make no gynecologic-oncology demand, adequacy, or supply-versus-demand claim; that cohort is carried solely as a supply-side replacement comparator.

We combined supply and demand in two ways. First, we computed urogynecologists per 100,000 women aged 65 years or older and the number of women with pelvic floor disorders per urogynecologist. Second, to move beyond a head count, we expressed supply as productivity-adjusted capacity by weighting each subspecialist by the clinical time devoted to urogynecologic care,[@temkin2022fte] so that a capacity-to-demand index (capacity relative to 2025 demand) reflects delivered care rather than the number of certified physicians; the urology-pathway contribution to that capacity is varied in sensitivity analysis. The supply trajectory carried the same entrant and empirically estimated departure inputs as the primary projection, extended from the fully modeled 2025-2029 horizon to 2050 to align with the demographic-demand horizon. Because that extension projects a 2016-2021 departure hazard and recent graduate output far beyond the calibrated window, the long-horizon supply-and-demand estimates are provisional and are reported to frame supply against demand, not as validated forecasts of clinical need. An interactive version of this supply-to-demand model, in which the productivity and pathway assumptions can be varied, accompanies the manuscript.

## Statistical Analysis

All analyses used R statistical software (version 4.4.2).[@rcoreteam2023] Workforce counts are reported as point estimates from the dynamic model accompanied by a 95% Monte Carlo parameter-uncertainty interval, which is conditional on the baseline workforce classification, the model structure, and the assumption of stable graduate output. Because the projection is a structural model of the inputs rather than an estimate from a random sample, no hypothesis tests were performed and no P values are reported.

# Results

## Baseline Workforce Characteristics (2025)

```{r table1}
workforce_summary <- load_workforce_data() %>% dplyr::filter(subspecialty_abbrev %in% c("GO", "URPS"))

table1 <- workforce_summary %>%
  transmute(
    Subspecialty = subspecialty,
    `Active Physicians (n)` = format(baseline_2025, big.mark = ",", trim = TRUE),
    `Model-implied weighted annual departure rate (%)` = sprintf("%.1f", annual_retirement_rate),
    `Mean Annual Departures (n)` = sprintf("%.1f", avg_annual_retirements),
    `Annual Graduates (n)` = annual_entrants
  )

kable(table1,
      caption = "**Table 1.** Estimated Active Workforce and Annual Stock-Flow Inputs. Annual graduates are the mean of the four most recent ACGME fellowship completion cohorts (AY2020-2021 through AY2023-2024; Gynecologic Oncology single-pathway, URPS both-pathway).",
      align = c("l", "r", "r", "r", "r"))
```

*The model-implied weighted annual departure rate is the `r primary_hazard_adjective()` age-band hazard applied to the 2025 active-age distribution; it is a model output, not a directly observed rate. The `r primary_hazard_adjective()` hazard is estimated from `r get_departure_events("GO")` Gynecologic Oncology and `r get_departure_events("URPS")` URPS departure events (`r get_person_years("GO")` and `r get_person_years("URPS")` person-years) over the fully observable 2016-2021 window; Appendix S6b gives the band-level event and person-year counts, and the ratio 95% intervals are reported in the Results. Mean annual departures are the dynamic model's four-year average and modestly exceed the active count times the 2025 rate because the cohort ages into higher-hazard bands over the horizon. Data derived from American Board of Obstetrics and Gynecology subspecialty certification records linked to National Provider Identifier and Medicare administrative files. URPS, Urogynecology and Reconstructive Pelvic Surgery; GO, Gynecologic Oncology.*

The 2025 active workforce comprised `r get_baseline("GO")` gynecologic oncologists and `r get_baseline("URPS")` both-pathway urogynecologists (Table 1). The `r primary_hazard_adjective()` age-band hazard implied weighted 2025 departure rates of `r get_annual_rate("GO")`% for Gynecologic Oncology and `r get_annual_rate("URPS")`% for URPS when applied to each cohort's active-age distribution. The two weighted rates are similar because both apply the same pooled age-band hazard, differing only through each cohort's age structure; the ordering of the two weighted rates was sensitive to pooling assumptions and should not be interpreted as evidence of a true difference in specialty-specific departure propensity. The hierarchical partial-pooled and fully unpooled hazards bracket these estimates (Appendix S6b).

Mean annual departures were `r get_avg_retirements("GO")` for Gynecologic Oncology and `r get_avg_retirements("URPS")` for URPS. Fellowship programs produced `r get_annual_entrants("GO")` Gynecologic Oncology and `r get_annual_entrants("URPS")` URPS graduates annually, above the corresponding mean annual departures.

## Four-Year Workforce Projections (2025-2029)

```{r table2}
.tramp <- readr::read_csv(here::here("data","graduation_active_transition_projection.csv"), show_col_types = FALSE)
table2 <- workforce_summary %>%
  dplyr::left_join(dplyr::select(.tramp, subspecialty_abbrev, projected_2029_ramped), by = "subspecialty_abbrev") %>%
  transmute(
    Subspecialty = subspecialty,
    `2025 Baseline (n)` = format(baseline_2025, big.mark = ",", trim = TRUE),
    `Projected 2029, immediate entry (n)` = format(round(projected_2029), big.mark = ",", trim = TRUE),
    `Projected 2029, transition-adjusted (n)` = format(round(projected_2029_ramped), big.mark = ",", trim = TRUE),
    `95% Parameter-Uncertainty Interval` = paste0(format(round(ci95_lower), big.mark = ",", trim = TRUE),
                      "-",
                      format(round(ci95_upper), big.mark = ",", trim = TRUE)),
    `Completion-to-Departure Ratio` = sprintf("%.2f", replacement_ratio)
  )

kable(table2,
      caption = "**Table 2.** Four-Year Workforce Projection and Fellowship-Completion-to-Departure Ratios (2025-2029), Dynamic Age-Structured Model.",
      align = c("l", "r", "r", "r", "r", "r"))
```

*Two projected 2029 counts are shown: the immediate-entry projection (graduates added to the active stock in their entry year, a structural upper estimate) and the transition-adjusted projection (applying the empirical certification-to-first-Medicare-activity entry ramp; Appendix S14), which defers part of the near-term growth. The 95% interval is the 2.5th to 97.5th percentile of a Monte Carlo parameter-uncertainty simulation and applies to the immediate-entry projection (partial: it omits baseline, roster, and model-form uncertainty; no separate interval is computed for the transition-adjusted point estimate). The fellowship-completion-to-departure ratio is annual fellowship completions divided by mean annual departures; because the numerator counts completions rather than verified active-practice entrants, it is a flow-balance measure, not a count of confirmed active replacements, and is unaffected by entry timing (above 1.05, at 0.95-1.05, below 0.95 flow balance) and does not measure adequacy relative to population demand. URPS is the both-pathway (ABOG and ABU) workforce. URPS, Urogynecology and Reconstructive Pelvic Surgery; GO, Gynecologic Oncology.*

```{r figure1, fig.cap="**Figure 2.** Projected active workforce, 2025-2029, under the status-quo entrant scenario (the recent four-year fellowship-graduate mean), applying the pooled age-band departure hazard to each cohort each year. Solid lines are the immediate-entry projection (each graduating cohort added to the active stock in its entry year), a structural upper estimate, with shaded 95% partial parameter-uncertainty bands (propagating age-band hazard and graduate-count uncertainty, but not baseline-classification, roster, or model-form uncertainty). Dashed lines (ending in hollow points) trace the lower transition-adjusted trajectory, from the 2025 baseline to the 2029 transition-adjusted count, which applies the empirical certification-to-first-Medicare-activity entry ramp and defers part of the near-term growth. Both cohorts grow through 2029; conservative (70% conversion) and optimistic (NRMP benchmark) entrant scenarios bracket these counts (Table 2). URPS, Urogynecology and Reconstructive Pelvic Surgery; GO, Gynecologic Oncology.", fig.width=7.2, fig.height=4.1}
fig_trajectory()
```

Annual fellowship graduate output exceeded projected departures in both cohorts, which are projected to grow through 2029 (Table 2; Figure 2). Both cohorts grew under all three entrant scenarios (Table 2). Under the status-quo immediate-entry scenario the projected 2029 workforce spanned, from the transition-adjusted lower bound to the immediate-entry upper bound, `r get_transition_ramped_2029("GO")` to `r get_projected("GO")` for Gynecologic Oncology (an increase of up to `r get_percent_change_magnitude("GO")`% over the `r get_baseline("GO")` baseline) and `r get_transition_ramped_2029("URPS")` to `r get_projected("URPS")` for the both-pathway URPS workforce (up to `r get_percent_change_magnitude("URPS")`% over `r get_baseline("URPS")`); both bounds exceed the 2025 baseline. The fellowship-completion-to-departure ratios were `r get_replacement_ratio("GO")` (Gynecologic Oncology) and `r get_replacement_ratio("URPS")` (URPS), that is roughly seven and five fellowship completions per projected annual departure, respectively, both well above one-for-one flow balance. Because the NRMP certified-position benchmark counts fellows beginning training, the optimistic scenario reaches the workforce only after the fellowship-duration lag and does not change the conclusion.

Every one-way sensitivity left both board-certified cohorts above replacement (Figure 3): even the full 2016-2023 window (which includes recent provisional departures with insufficient follow-up and may be upward biased) kept them well above replacement (Appendix S12), and a departure rate up to three times the empirical estimate did not move either below replacement (one-way minimum ratio `r get_grid_oneway_min("URPS")`). In the full multidimensional grid crossing the observation window, a departure-rate multiplier up to threefold, and graduate practice-conversion down to 70% (`r get_grid_total("GO")` cells per cohort), Gynecologic Oncology was above replacement in `r get_grid_above("GO")` of `r get_grid_total("GO")` cells (`r get_grid_at("GO")` at replacement, `r get_grid_below("GO")` below) and URPS in `r get_grid_above("URPS")` (`r get_grid_at("URPS")` at replacement, `r get_grid_below("URPS")` below); at the joint extreme of the provisional full window, tripled departures, and 70% conversion, Gynecologic Oncology was at replacement (`r get_grid_worst("GO")`) and URPS was below replacement (`r get_grid_worst("URPS")`) (Appendix S12). Varying the required inactivity duration rather than the calendar window gave the same conclusion: re-deriving departures directly from Medicare Part B activity at a 2-, 3-, or 4-year inactivity threshold (a more aggressive definition than the multi-source primary), both cohorts remained well above replacement at every threshold (Gynecologic Oncology `r get_inactivity_ratio(2,"GO")` at 2-3 years rising to `r get_inactivity_ratio(4,"GO")` at 4 years; URPS `r get_inactivity_ratio(2,"URPS")` rising to `r get_inactivity_ratio(4,"URPS")`), with a stricter threshold raising the ratio because fewer exits meet the longer absence requirement (Appendix S27). The primary `r primary_hazard_adjective()` age-band hazard (Gynecologic Oncology `r get_primary_ratio("GO")`, URPS `r get_primary_ratio("URPS")`; 95% intervals `r get_primary_ci("GO")` and `r get_primary_ci("URPS")`, both above replacement) is bracketed by the hierarchical partial-pooled and fully unpooled hazards, reported as sensitivities: Gynecologic Oncology `r get_hier_ratio("partial_pooled","GO")` partial-pooled and `r get_hier_ratio("unpooled","GO")` unpooled, URPS `r get_hier_ratio("partial_pooled","URPS")` and `r get_hier_ratio("unpooled","URPS")` (Appendix S6b, Appendix Table S28). Folding ABU-hazard uncertainty (0.5 to 2 times) into that hierarchical envelope left it above replacement (`r get_hier_ratio("partial_pooled_abu_2x","URPS")` to `r get_hier_ratio("partial_pooled_abu_0.5x","URPS")`) (Appendix Table S28). Beyond the flat four-year mean, the conclusion also held across distinct graduate-supply assumptions (near-term cohort accounting, that is the most recent completing class; a contraction to the recent annual low; and a cautious extrapolated trend), with both cohorts remaining above replacement (Gynecologic Oncology `r get_grad_scenario_ratio("contraction","GO")` under contraction to `r get_grad_scenario_ratio("cohort_accounting","GO")` under cohort accounting; URPS `r get_grad_scenario_ratio("contraction","URPS")` to `r get_grad_scenario_ratio("cohort_accounting","URPS")`) (Appendix S29).

```{r figure2, fig.cap="**Figure 3.** Fellowship-completion-to-departure ratio for each cohort across the primary analysis and selected sensitivity analyses. The dashed line marks one-for-one replacement; the shaded region is below replacement. Points right of the line indicate that annual fellowship completions exceed classified departures. Sensitivities span the anchored-definition baseline (Appendix S26), the Open Payments-inclusive definition (Appendix S15), adding half or all expected age-specific deaths (Appendix S25), the provisional full 2016-2023 window, a tripled departure rate (one-way minimum), and the joint worst-case grid cell (full window plus tripled rate plus 70% graduate conversion; Appendix S12). Only the joint worst case reaches or dips to replacement. URPS, Urogynecology and Reconstructive Pelvic Surgery; GO, Gynecologic Oncology.", fig.width=7.4, fig.height=5.0}
fig_robustness()
```

## Sensitivity Analyses

```{r age-departure-prose, results='asis'}
.win <- readr::read_csv(here::here("data", "departure_window_sensitivity.csv"),
                        show_col_types = FALSE)
g <- function(df, ab, col, lbl=NULL) { d <- if(is.null(lbl)) df else df[df$label==lbl,]; d[[col]][d$subspecialty_abbrev == ab] }
cat("Estimated age at classified departure is younger than the median age of about 64 years at which obstetrician-gynecologists customarily retire from clinical practice.[@rayburn2015retirement; @tucker2025retirement] Because voluntary retirement is only one route out of practice, distinct from forced (health, competency, or legal) exits, early retirement, and non-clinical career change,[@tucker2025retirement] the younger ages may reflect nonretirement workforce exits, but until the physician-level adjudication is complete may also reflect classification or administrative-ascertainment error. Because age is reconstructed from certification year and the departure endpoint remains provisional, and because a young estimated median age is sensitive to misclassification, the per-subspecialty age-at-departure estimates are reported in the Supplemental Digital Content (Appendix S3) rather than in the main text.\n\n")
cat(sprintf(
  "Passing each observation window through the same dynamic model, the completion-to-departure ratio for Gynecologic Oncology was %.2f in the primary fully-observable window (2016-2021), %.2f dropping two further years (2016-2019), and %.2f in the full window (2016-2023, which includes recent provisional departures with insufficient follow-up); for URPS the corresponding ratios were %.2f, %.2f, and %.2f. Both subspecialties remained above replacement in every window, and each stayed above replacement when the departure rate was doubled.",
  g(.win, "GO", "dynamic_ratio", "fully_obs"), g(.win, "GO", "dynamic_ratio", "drop2"), g(.win, "GO", "dynamic_ratio", "full"),
  g(.win, "URPS", "dynamic_ratio", "fully_obs"), g(.win, "URPS", "dynamic_ratio", "drop2"), g(.win, "URPS", "dynamic_ratio", "full")))
```

No single parameter reached replacement at a plausible value: the departure rate would have to rise `r get_breakeven_rate_range()` or graduate output fall `r get_breakeven_drop_range()` (Appendix S10). Below-replacement estimates required combining adverse assumptions, and their form depended on the window: in the primary fully-observable window only a tripled rate, 70% entry, and a 0.65 entrant-effort multiplier together fell below one, whereas in the provisional full 2016-2023 window (which may over-state departures) a tripled rate and 70% entry alone sufficed (Appendix S12).

## Supply Relative to Demand

Set against the population it serves, projected urogynecology supply outpaced demographic demand across the 2025-2050 horizon (accompanying supply-to-demand model). The population of United States women aged 65 years or older, in whom pelvic floor disorders concentrate, was projected to rise about 29% (from 34.7 to 44.8 million) and the population living with a pelvic floor disorder about 19% (from 33.1 to 39.5 million), while the projected workforce grew faster, by about 79% (index 179, 2025 = 100). As a result, urogynecologists per 100,000 women aged 65 years or older rose from 3.9 in 2025 to about 5.4 in 2050, and the number of women with a pelvic floor disorder per urogynecologist fell from roughly 24,800 to 16,500. Expressed as productivity-adjusted capacity rather than head count, the capacity-to-demand index rose above its 2025 value across the horizon and remained above one under the base-case urology-pathway contribution; because supply grew from a larger proportional base than demand, that ordering was preserved across the plausible range of productivity and pathway assumptions examined in sensitivity analysis. These supply-and-demand estimates are national aggregates under stable prevalence and care-seeking and rest on a long-horizon extrapolation of the departure hazard; they indicate whether the number and capacity of urogynecologists keeps pace with a growing older-female population, not whether that capacity reaches the markets or procedures where demand is concentrated.

## Operative-Workforce Analysis (Exploratory)

Medicare fee-for-service-defined loss of qualifying surgery was strongly dependent on follow-up duration and could not be validated against available administrative signals: most apparent cessations were not accompanied by complete clinical-practice departure, and because a true operative retiree may keep billing clinic visits, the signal could not distinguish genuine cessation of surgery from loss of Medicare visibility. Operative-workforce replacement was therefore not estimated; operator prevalence, cessation rates, time-to-event operative entry, concordance, and volume-weighted calculations are reported only in the Supplemental Digital Content (Appendix S11).

# Discussion

## Principal Findings

Using primary-sourced inputs (active counts, actual ACGME fellowship graduate counts, and `r primary_hazard_adjective()` age-band departure hazards), we found that annual fellowship graduate output exceeds administratively identified departures in both board-certified subspecialties, with fellowship-completion-to-departure ratios of `r get_replacement_ratio("GO")` for Gynecologic Oncology and `r get_replacement_ratio("URPS")` for the both-pathway URPS workforce, both well above one-for-one flow balance. Under this administrative model, national fellowship completion does not appear constrained by near-term aggregate headcount replacement requirements over the four-year horizon; longer-term sustainability through the larger retirement wave, and whether the pipeline binds in particular regions, institutions, or operative niches, are beyond this near-term national projection.

This conclusion turns on three inputs that prior estimates have often specified differently: the number of physicians actually completing training each year, measured from ACGME graduate counts rather than assumed from program capacity or from match entry (which counts fellows beginning, not finishing, and runs ahead of graduate output in a growing field); the size of the active, rather than merely certified, workforce, which is materially smaller than the certified count; and the departure rate, estimated from the study cohort's own observed departures over a right-censoring-safe window (a `r primary_hazard_adjective()` gynecologic-subspecialty age-band hazard applied to each subspecialty's age distribution) rather than from general-physician age-band assumptions.

## Headcount Versus Operative-Capacity Replacement

Whether head-count replacement implies replacement of operative capacity is a distinct question our data could not resolve. An exploratory Medicare fee-for-service analysis found that most apparent operative cessations were not accompanied by complete clinical-practice departure (only `r get_op_corroborated("GO")`% to `r get_op_corroborated("URPS")`% coincided with an independent departure signal), and estimates of graduate operative entry depended heavily on follow-up length; because a true operative retiree may keep billing clinic visits, these signals could not separate genuine cessation of surgery from loss of Medicare visibility.[@poushay2018; @conn2018; @hedden2017] Because Medicare fee-for-service omits the commercial and facility-billed surgery that constitutes much of these subspecialties' operative volume, operative-capacity replacement remains unresolved and will require all-payer or institutional operative-log data. The conceptual point stands: because a surgeon can leave operative practice while remaining clinically active, adequate head count does not guarantee adequate surgical capacity, and this distinction should be measured rather than assumed.

## Comparison with Existing Literature

Our findings extend the most recent gynecologic oncology fellowship analysis: Holtzman and colleagues documented programs rising from 38 to 65 and annual positions from 36 to 80 between 2004 and 2023, alongside geographic maldistribution;[@holtzman2025] our completion-to-departure ratio of `r get_replacement_ratio("GO")` quantifies the inflow-outflow imbalance underlying that observation by measuring actual graduate output against empirically estimated departures. Earlier need-based projections did not measure graduate output,[@wallace2010] which we find substantially exceeds program-capacity assumptions, so documented worsening access to gynecologic oncologists[@desjardins2023] cannot be attributed to near-term national head-count replacement failure alone. For urogynecology, the single prior national supply-and-demand projection reported roughly 27,650 women with a pelvic floor disorder per subspecialist by 2045, establishing that a rising head count does not erase the disease burden.[@brueseke2016] We advance that analysis by estimating departure empirically from six administrative streams with quantified uncertainty rather than assuming fixed-age retirement; counting an active both-pathway roster validated by Medicare billing rather than a certification or society-membership head count that can undercount the urology pathway;[@wallace2010] expressing supply as productivity-adjusted capacity, since clinical capacity, not the number of bodies, is what patients experience;[@stitzenberg2015; @temkin2022fte] and reporting per-capita coverage rather than a single ratio. Because the demand denominator (women with any modeled pelvic floor disorder) is an upper-bound epidemiologic construct rather than a count of visits or operations,[@nygaard2008] we present coverage as relative and do not interpret it as the proportion of clinical need met.

Our departure estimates align with the largest national attrition analyses, compared at the correct level. Elemosho and colleagues, using the same Medicare and NPPES data, reported an all-surgeon rate of 1.5% to 1.7% per year and placed gynecologic oncology among the lowest-attrition specialties (adjusted hazard ratio 0.73 versus general surgery);[@elemosho2026] the weighted gynecologic oncology rate implied by the pooled hazard (`r get_annual_rate("GO")`%) is numerically lower, though this descriptive comparison is limited by the differing departure definition and pooled-hazard modeling and is not independent confirmation of the external hazard ratio. Urogynecology was subsumed in a generalist obstetrics-gynecology category with high attrition (hazard ratio 2.23), but that reflects a largely non-Medicare, reproductive-age panel and should not be attributed to the surgical subspecialists; in the largest study to date surgical specialties were among the lowest-attrition groups and elevated generalist attrition fell once the older Medicare panel was accounted for (1.62 to 0.79).[@rotenstein2025] A dedicated oncology analysis found gynecologic oncology attrition steeply experience-graded and predominantly late-career,[@tuia2023] supporting our age-stratified approach. These studies quantified surgeon outflow but did not test whether fellowship output replaces it, the question we answer.

Prior projections also show the hazard of assuming a departure rate: an emergency-medicine surplus collapsed when measured attrition exceeded the assumed 3%,[@marco2021; @gettel2022] and a transplant-surgery projection based on retirement alone missed early-career attrition.[@delman2023] We therefore derived the rate empirically over a right-censoring-safe window and confirmed both board-certified subspecialties stay above replacement across every window and at twice the rate.

## Implications for Policy and Training

### Interpreting Replacement Balance

Under the modeled assumptions, we found no evidence of a near-term national aggregate headcount replacement gap that would, by itself, require expanding fellowship positions over 2025-2029; this does not imply the workforce meets population demand. Documented access limitations cannot be attributed to near-term national headcount replacement failure alone: the share of U.S. counties with no obstetrician-gynecologist was essentially unchanged at about 47% between 2010 and 2020-2021 despite a stable-to-growing supply,[@chen2026geographic] and retention is worse in socioeconomically deprived areas,[@lee2024] consistent with distribution being a lever for access; because this study contains no subspecialty-specific wait-time, accessibility, or demand data, it cannot establish maldistribution as the principal cause.

Our supply-side finding also does not contradict federal projections of a worsening aggregate obstetrics-gynecology shortage: HRSA-based modeling projects the pooled obstetrician-gynecologist workforce to fall about 9% in full-time-equivalent supply by 2035, with only six states adequate and nonmetropolitan adequacy near 51%.[@rayburn2026shortage] Those projections model full-time-equivalent supply against population demand for the combined workforce, which the federal model cannot separate into generalists and subspecialists,[@rayburn2025complexities] whereas we model subspecialty headcount inflow against outflow over a nearer horizon. A subspecialty can be above headcount replacement yet belong to a workforce that is below demand adequacy and geographically maldistributed, particularly because gynecologic surgical subspecialists concentrate in metropolitan areas.

## Strengths and Departure-Detection Methodology

This analysis combines primary-source workforce and fellowship data with explicit administrative-classification and modeling assumptions: an active workforce from certification records linked to administrative data, ACGME graduate counts, and departure rates estimated empirically over a right-censoring-safe window, matching a national surgeon-attrition benchmark from the same data.[@elemosho2026] The multi-source classifier integrates certification, industry-payment, prescribing, enrollment, and claims signals rather than a single source such as Medicare billing cessation, intended to reduce reliance on any single source and the associated risk of misclassifying a low-volume subspecialist who stops billing Medicare but keeps operating. Back-testing reproduced observed active head count within about 1% (Appendix S3) but does not validate individual classifications; the classifier agreed with state licensure on active status (negative predictive value 0.98) yet flagged more departures than licensure recorded (about 11% versus 3%; Cohen kappa 0.06), as expected because physicians retain licenses after leaving practice. Because cessation of Open Payments reporting (the sole firing source for about 30% of detected departures) is not a validated indicator of leaving practice, the primary definition requires a corroborating non-Open-Payments anchor (Medicare Part B or Part D cessation, NPPES deactivation, or ABMS lapse); under the Open Payments-inclusive definition the ratios were lower (about `r get_opsens_ratio("op_inclusive","GO")` for Gynecologic Oncology and `r get_opsens_ratio("op_inclusive","URPS")` for URPS) and under the primary definition higher, but both cohorts remained above replacement under every source-inclusion rule (Appendix S15). Because the primary outcome is departure, not active status, a high negative predictive value does not establish positive predictive value or sensitivity, and automated corroboration cannot fill the gap: within the Open-Payments-inclusive layer a non-Open-Payments signal supported `r get_ppv_proxy("GO")`% (Gynecologic Oncology) and `r get_ppv_proxy("URPS")`% (URPS) of detected departures (Appendix S7b), but the anchored endpoint retains only events carrying such an anchor, so 100% are corroborated by construction, a tautology rather than validation. Genuine individual-level validation requires the prespecified two-reviewer chart adjudication, complete but pending (Appendix S7); until it is complete we do not claim the departure classifier is individually validated and report the endpoint as provisional. Death was under-ascertained by both systems; missed deaths leave deceased physicians in the active stock and drop true departure events from the numerator, biasing the departure rate downward and the completion-to-departure ratio upward. To bound this we applied an age- and sex-specific expected-mortality sensitivity, adding expected deaths to the departure numerator (SSA Period Life Table, 2020; Appendix S25), a conservative upper bound because general-population mortality overstates physician mortality. Even so the ratios fell only to `r get_mortality_ratio_all("GO")` for Gynecologic Oncology and `r get_mortality_ratio_all("URPS")` for URPS (`r get_mortality_ratio_half("GO")` and `r get_mortality_ratio_half("URPS")` if half were missed), both above replacement.

## Limitations

Several limitations warrant acknowledgment. First and most important, this analysis addresses only headcount replacement, not whether the workforce's size, distribution, or clinical capacity meets population demand; and the demand model is specific to urogynecology (pelvic floor disorder prevalence), so gynecologic oncology is reported purely as a supply-side comparator, with no gynecologic-oncology demand or adequacy inferred. We count physicians, not full-time equivalents; because obstetrician-gynecologists increasingly work part-time and the workforce is feminizing, headcount may overstate effective clinical supply.[@rayburn2025complexities] An exploratory clinical-effort sensitivity leaves the completion-to-departure ratio unchanged when the same factor applies to incumbents and graduates, but the ratio would fall if new graduates delivered less direct clinical and operative capacity than those they replace (more administrative and part-time practice, childbearing interruptions, lower operative volume); this is an effective-capacity proxy rather than headcount replacement, and an exploratory entrant-capacity sensitivity remained above a one-for-one effective-capacity proxy even at a 0.65 entrant-to-incumbent multiplier, a 35% reduction, though no evidence establishes 0.65 as the appropriate lower bound (Appendix S9).[@temkin2022fte; @benedetti2004productivity] A distinct forward risk, beyond this horizon, is that younger cohorts anticipate earlier retirement.[@muzyka2025generational]

Second, the model holds hazards and graduate output constant and does not model secular trends; national attrition rose about 40% between 2013 and 2019,[@rotenstein2025] though the conclusion is robust to a departure rate twice the empirical estimate. Graduate output is likewise held constant while fellowship applicant demand has narrowed toward the number of positions.[@humphries2026] In the 2025 Match (for 2026 appointments), 99 applicants preferred gynecologic oncology for 88 positions (one unfilled), and although aggregate URPS applicant demand exceeded positions, the pathways diverged: the urology track offered 18 positions to 17 applicants who preferred it and filled 14, whereas the obstetrics-gynecology track filled all its positions.[@nrmp2025] Sustained output is therefore not guaranteed; reaching replacement would nonetheless require graduate output to fall `r get_breakeven_drop_range()` (Appendix S10), substantially larger than the applicant-position gaps observed in the most recent Match, and continued growth depends disproportionately on obstetrics-gynecology applicant interest and improved urology-pathway recruitment. The Monte Carlo interval propagates hazard and graduate-count uncertainty but not baseline classification error, so it is not a total-uncertainty interval.

Third, the primary model adds graduates to the active stock in their entry year, whereas real entry has a lag. We estimated a certification-to-first-Medicare-activity transition empirically by linking board-certification cohorts (2013-2019) to their first observed Medicare activity (Appendix S14); this proxies, but does not equal, the fellowship-completion-to-active-practice lag, because certification is not graduation and Medicare visibility is not all practice. Among certificants eventually observed in Medicare, `r get_transition_kyr("GO", 3)`% appeared within three years, though only `r get_transition_ceiling("GO")`% (Gynecologic Oncology) and `r get_transition_ceiling("URPS")`% (URPS) were ever Medicare-visible. Applying the timing ramp defers `r get_transition_deferred("GO")`% (Gynecologic Oncology) and `r get_transition_deferred("URPS")`% (URPS) of the projected four-year growth beyond 2029 (transition-adjusted 2029 counts `r get_transition_ramped_2029("GO")` and `r get_transition_ramped_2029("URPS")`), but the workforce still grows and the replacement balance, a flow ratio, is unchanged. The `r 100-get_transition_ceiling("GO")`% to `r 100-get_transition_ceiling("URPS")`% never observed in Medicare plausibly reflect non-Medicare practice, though some may not enter U.S. subspecialty practice; the 70% to 100% conversion sensitivity brackets any true non-entry. A further estimated 10% to 15% of graduates later pursue non-clinical careers or leave U.S. practice.[@bakkensen2023; @knell2023] Even at 70% conversion the conclusion holds.

Fourth, the active baseline comes from a multi-source classification checked for licensure concordance but not yet individually adjudicated (Appendix S7); because several signals are Medicare-based, the classifier is less sensitive for younger, non-Medicare panels, which is why we excluded maternal-fetal medicine and reproductive endocrinology. The active-baseline gate and the anchored historical departure-event definition were developed for different analytic purposes and are not identical: the baseline uses the multi-source cohorting classification, whereas the outflow events require a non-Open-Payments anchor. In the prespecified consistent-definition baseline sensitivity, which regenerates the 2025 active stock under the same anchored departure rule (Appendix S26), the anchored stock rose modestly to `r get_consistent_baseline("GO")` for Gynecologic Oncology and `r get_consistent_baseline("URPS")` for URPS, and the completion-to-departure ratios remained well above replacement (`r get_consistent_ratio("GO")` and `r get_consistent_ratio("URPS")`), confirming that pairing a broad baseline with an anchored departure definition does not materially bias the stock-flow balance. The estimated 2025 baseline is carried forward from administrative data ending 2023-2024, but it is well supported: about `r get_baseline_directly_pct("GO")`% of the Part B-observable Gynecologic Oncology stock and `r get_baseline_directly_pct("URPS")`% of the ABOG-pathway URPS stock had Medicare Part B activity in the latest administrative year (2023), and only `r get_baseline_carried3("GO")` and `r get_baseline_carried3("URPS")` physicians respectively rested on a three-year-or-longer carry-forward; removing those long-lag physicians as the most adverse baseline-lag assumption did not weaken the conclusion (ratios `r get_baseline_lag_ratio("GO")` and `r get_baseline_lag_ratio("URPS")`, slightly above the primary) (Appendix S30).

Fifth, age was estimated from certification year; it correlated strongly with external sources (Healthgrades r = 0.91, graduation-year r = 0.96; median absolute error 2 years, Appendix Table S6), and because the same estimate sets both the hazards and the active-age distribution the small under-estimate largely cancels; shifting all ages up 2 to 4 years modestly lowered the ratio but left both cohorts well above replacement (Appendix Table S24).

Sixth, the workforce was modeled as a closed system; re-entry, internationally trained subspecialists, and emigration are not represented, and administrative lag leaves the most recent cohorts provisional. Interstate relocation does not change national totals.[@bond2023]

Seventh, the both-pathway URPS baseline adds ABU-certified urogynecologists from an external roster, dated by certification year and assigned the ABOG-URPS hazard because ABU-pathway departure hazards are not independently estimable in this cohort. In a sensitivity that modeled the ABU component separately (a deterministic approximation, distinct from the hierarchical ABU-hazard scaling of Appendix Table S28) and scaled its hazard from half to twice the ABOG rate, URPS remained well above replacement (ratio `r get_abu_pathway_range()`; Appendix S11.5), and this both-pathway design removes the larger stock-flow mismatch of counting one pathway's stock against graduates from both.[@plaska2025]

Eighth, these subspecialties are feminizing, and women leave clinical practice at higher rates and younger ages;[@rotensteinsex2026; @romero2025feminization] in a sex-stratified sensitivity drawing entrants from the roughly 85%-female obstetrics-gynecology resident pool (an ABOG-pathway URPS sensitivity; the urology URPS pathway is not sex-modeled and the assumption does not apply to it), all cohorts remained above replacement with ratios essentially unchanged, because entrants are young and depart at low rates regardless of sex (Appendix S8). The early-career feminization effect would emerge only over a longer horizon.

# Conclusion

Under the study's assumptions, current fellowship completion substantially exceeds near-term administratively classified clinical-practice departures among board-certified gynecologic oncologists and urogynecologists: Gynecologic Oncology (completion-to-departure ratio `r get_replacement_ratio("GO")`) and the both-pathway Urogynecology and Reconstructive Pelvic Surgery workforce (`r get_replacement_ratio("URPS")`). Under this provisional administrative model, near-term national headcount balance alone does not provide evidence that additional fellowship positions are required solely to offset departures through 2029; this four-year horizon does not address longer-term pipeline sustainability through the larger retirement wave, geographic access, or operative capacity, for which fellowship-capacity decisions may still warrant separate consideration. The headcount conclusion is unusually robust, reversing only under a joint extreme that combined a provisional observation window, a tripled departure rate, and 70% graduate conversion. More broadly, measuring actual graduate output against administratively observed departures offers a reusable empirical test of workforce-cliff claims that have more often been asserted from demographic aging than measured, and the same approach could be applied to other surgical subspecialties facing similar concern.

These conclusions concern national headcount balance only and do not establish workforce adequacy relative to population demand, geographic access, clinical full-time equivalents, or operative capacity; Medicare fee-for-service data in particular could not distinguish true cessation of operative practice from loss of claims visibility, so operative-workforce replacement remains unresolved and requires all-payer or institutional operative data.

# References

<div id="refs"></div>
